Decompositions of congruence subgroups of Chevalley groups

### 2015-11-08 paper.tex
More fixes

\title [Decompositions of congruence subgroups of Chevalley groups] {Decompositions of congruence subgroups \\ of Chevalley groups}
\keywords {parabolic decomposition, bounded generation, $\SL_2$-factorization, subsystem factorization. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author{Sergey Sinchuk}
\author{Andrei Smolensky}
\email{sinchukss {\it at} yandex.ru; andrei.smolensky {\it at} gmail.com}
\thanks{Authors of the present paper acknowledge the financial support from Russian Science Foundation grant 14-11-00297}

@@

The next main result is the relative Dennis---Vaserstein decomposition, which holds for a broader class of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$. Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\asr(I) \leqslant d$ for $\Phi=\rF_4, \rE_\ell$ and $\{r, s\} = \{1, \ell \}$, 
 \item $\sr(I) \leqslant d$, in all other cases.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

@@

\textsc{Case $\Phi=\rD_\ell$.} From the proof of \cref{item:asrUnipD} it follows that there exists $h_1 \in \EP_\ell(R, I)$ such that for $v'=h_1\cdot v$ one has $v'_2=v'_1-1\in I$.
Clearly, for $v'' = z_{-\alpha\ssub{2}}(-v'_2, 1)\cdot v'$ one has $v''_1=1$, hence by \cref{lemma:Chevalley-Matsumoto} there exists $h_2 \in \U(\Phi^-, I)$ such that the element
$g'=h_2 \cdot z_{-\alpha\ssub{2}}(-v'_2, 1) \cdot h_1 \cdot g$ fixes $v^+$. 
Using the Levi decomposition we can write $g'=h \cdot y \cdot x \cdot g$ for some $y\in\U(\Sigma^-_\ell, I)$, $x \in \U(\Sigma^+_\ell, I)$, $h\in\E(\Delta_\ell, R, I)$.
It is clear that $x$, $y$ are the desired elements.
\end{proof}s

@@

\begin{proof}[Proof of \cref{thm:DennisVaserstein}]
First of all, notice that in the case $\Phi=\rG_2$ the assertion of the theorem formally follows from \cref{thm:srRI1} and identities~\ref{rel:h-comm}--\ref{rel:h-w}.
Since the equality $A_{rs} = \E(\Phi, R, I)$ implies $A_{sr} = \E(\Phi, R, I)$, it suffices to consider only the following possibilities for $\Phi$, $s$, $r$:
\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                  & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                  & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1  \\
\rF_4                                & 4    &1     & 26             & \text{minimal}       & \rC_3                     & 6              & 3\end{array}\]
 \caption{List of the cases considered in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}

>>>>

\title [Decompositions of congruence subgroups of Chevalley groups] {Decompositions of congruence subgroups \\ of Chevalley groups}
\keywords {parabolic decomposition, bounded generation, $\SL_2$-factorization, product decomposition. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author{Sergey Sinchuk}
\author{Andrei Smolensky}
\email{sinchukss {\it at} yandex.ru; andrei.smolensky {\it at} gmail.com}
\thanks{Authors of the present paper acknowledge the financial support from Russian Science Foundation grant 14-11-00297}

@@

The next main result is the relative Dennis---Vaserstein decomposition, which holds for a broader class of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system of rank $\ell\geq 2$ and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$. Denote the distance between $r$ and $s$ by $d$.
Assume that $\Phi$, $I$ and $\{r, s\}$ satisfy either of the following assumptions:
\begin{thmlist}
 \item $\sr(I) \leqslant d$ for classical $\Phi$,
 \item $\sr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{ 2, \ell \}$.
 \item $\asr(I) \leqslant d$ for $\Phi=\rE_\ell$ with $\{r, s\} = \{1, \ell \}$,
 \item $\asr(I) \leqslant d$ for $\Phi=\rF_4$.
\end{thmlist}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5 and~4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

@@

\textsc{Case $\Phi=\rD_\ell$.} From the proof of \cref{item:asrUnipD} it follows that there exists $h_1 \in \EP_\ell(R, I)$ such that for $v'=h_1\cdot v$ one has $v'_2=v'_1-1\in I$.
Clearly, for $v'' = z_{-\alpha\ssub{2}}(-v'_2, 1)\cdot v'$ one has $v''_1=1$, hence by \cref{lemma:Chevalley-Matsumoto} there exists $h_2 \in \U(\Phi^-, I)$ such that the element
$g'=h_2 \cdot z_{-\alpha\ssub{2}}(-v'_2, 1) \cdot h_1 \cdot g$ fixes $v^+$. 
Using the Levi decomposition we can write $g'=h \cdot y \cdot x \cdot g$ for some $y\in\U(\Sigma^-_\ell, I)$, $x \in \U(\Sigma^+_\ell, I)$, $h\in\E(\Delta_\ell, R, I)$.
It is clear that $x$, $y$ are the desired elements.
\end{proof}

@@

\begin{proof}[Proof of \cref{thm:DennisVaserstein}]
%First of all, notice that in the case $\Phi=\rG_2$ the assertion of the theorem formally follows from \cref{thm:srRI1} and identities~\ref{rel:h-comm}--\ref{rel:h-w}.
Since the equality $A_{rs} = \E(\Phi, R, I)$ implies $A_{sr} = \E(\Phi, R, I)$, it suffices to consider only the following possibilities for $\Phi$, $s$, $r$:
\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & s    &r      & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell, \ \ell\geqslant 2           & 1    &\ell   & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell, \ \ell\geqslant 2           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell, \ \ell\geqslant 4           & 1    &\ell   & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell, \ \ell\geqslant 4           & \ell &\ell-1 & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell, \ \ell=6, 7                  & \ell &2      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell, \ \ell=6, 7                  & \ell &1      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1  \\
\rF_4                                & 4    &1     & 26             & \text{minimal}       & \rC_3                     & 6              & 3\end{array}\]
 \caption{List of the cases considered in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}
